%----------------------------------------------------------------------------------------
%	PREFACE PAGE
%----------------------------------------------------------------------------------------

\section*{PREFACE}
\addcontentsline{toc}{chapter}{\protect\numberline{}PREFACE}

\par You are about to meet a computer that feels out of place for its time.  It
is slower, larger, and more expensive than most computers of its era.  In many
ways, it is a technological anachronism designed for a niche market of
hobbyists and enthusiasts.  But that's not all.

\medskip\par
The Commander X16 reaches into the past and brings back many things that
were lost.  It is a computer that recaptures the "soul" of the early days of
home computing.  Together with this manual, anyone should be able to sit down
and begin to explore computing with all the modern layers of abstraction
stripped away.  No prior knowledge of computing or even typing should be
required in order to start your journey into the world of computers!

\medskip\par
This manual should be readable by experts and novices alike.  Even children as
young as 6 or 7 should be able to follow along with the step-by-step examples.
There is no reason you need to read this manual in order.  After reading
Chapter 1 (Getting Started) you can go directly to a chapter that interests you
and start reading.  The first page of each chapter contains a small sample
program to type in.  Type it exactly for a demonstration of what you will learn
in that chapter.  The rest of the chapter will explain the details as you read,
and contain more sample programs to try out.

\medskip \par
The Commander X16 was created with the intention that you can simply turn it on
and start learning, doing, and creating.  The friendly blue screen and colorful
butterfly logo invite you to start typing BASIC commands and programs.  The
built-in SD card reader allows you to save your work as well as load the
programs and art that others create, all without needing to troubleshoot
expensive antique disk drives and data sets...although it supports those too!

\medskip\par
Computers have become an important part of our everyday lives, and yet most
people never delve past the surface of the user interfaces presented to them.
Children are given touch screen devices before they can even talk, and adults
go through their entire lives without learning about the magic that happens
behind their screens.  The Commander X16 reverses this trend by putting the
user back in control of a computer that they can fully understand.

\medskip\par
The Commander X16 is the perfect computer for this day and age!
